Linear stability of magnetized massive protoplanetary disks

### 2014-03-31 paper/results_azi.tex
tex updates

\subsection{Ideal disks without GI} 
Here we consider an isothermal disk with $Q=0.2$
($Q_\mathrm{2D}=0.72$) and $\beta=10$ in the limit of ideal MHD
($\Lambda_0=100$). Gravitational instability is not expected because 
Fig. \ref{compare_growth3} shows that even for $Q=0.18$, GI is 
surpressed for $\beta \lesssim 15$. We consider azimuthal field
strengths with $\epsilon=[0,3]$.

Fig. shows the MRI growth rates for the above system as a function of
$k_x$. The top panel are modes with even midplane symmetry: 
$W^\prime(0)\sim 0$ for $k_x\to0$, and the bottom panel are modes with odd midplane 
symmetry: $W(0)\sim 0$ for $k_x\to0$. The latter set of modes were 
excluded in the previous sections by midplane boundary 
conditions. The plot also include growth rates computed in the Cowling
approximation. As expected, none of the modes are energetically
dominated by self-gravity, $\avg{E_g}<\avg{E_m}$.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm 0cm]{figures/compare_growth3_tilted_even.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm 0.52cm]{figures/compare_growth3_tilted_odd.ps}
  \caption{Growth rates in massive isothermal disks with $Q=~0.2$ and
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation, while
    the solid curves are computed from the Cowling approximation. The
    top panel correspond to even modes and the
    bottom panel correspond to odd modes, which have
    $W^\prime(0)\sim0$ and $W(0)\sim0$, respectively, for $k_x\to0$.
    \label{compare_growth3_tilted}}
\end{figure}

@@

Consider first the even modes in the top panel of
Fig. \ref{compare_growth3_tilted}. Self-gravity slightly stabilizes
modes with small $k_x$ as $B_y$ is increased. SG destabilizes modes
with  $k_xH\gtrsim O(1)$, and is most effective for purely
vertical fields: for $\epsilon=0,\, k_xH\simeq 1.4$, SG increases the
growth rate by $\sim 30\%$. Consequently the cut-off wavenumber is
larger when SG is included. This destabilization effect weakens with
$B_y$ because the total magnetic pressure increases, which generally
opposes GI, although the modes are fundamentally associated with MRI.

The odd modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. Self-gravity has negligible effect when 
the field is purely vertical ($B_y=0$). SG stabilizes all wavelengths,
and is most effective with increasing $B_y$ and $k_xH =  O(1)$.

%With $B_y\neq 0$, the MRI is compressible even in the Cowling
%approximation \citep{pessah05}.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0cm]{figures/result_tilted_cowling.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.cm]{figures/result_tilted_fullsg.ps} 
  \caption{caption
    \label{result_tilted}}
\end{figure}

\subsection{Resistive disks with GI}

>>>>

\subsection{Ideal disks with MRI} 
Here we consider an isothermal disk with $Q=0.2$
($Q_\mathrm{2D}=0.72$) and $\beta=10$ in the limit of ideal MHD
($\Lambda_0=100$). Gravitational instability is not expected because 
Fig. \ref{compare_growth3} shows that even for $Q=0.18$, GI is 
surpressed for $\beta \lesssim 15$. We consider azimuthal field
strengths with $\epsilon\in[0,3]$.

Fig. shows the MRI growth rates for the above system as a function of
$k_x$. The top panel are modes with even midplane symmetry: 
$W^\prime(0)= 0$ for $\epsilon\to0$, and the bottom panel are modes with odd midplane 
symmetry: $W(0) = 0$ for $\epsilon\to0$. The latter set of modes were 
excluded in the previous sections by midplane boundary 
conditions. The plot also include growth rates computed in the Cowling
approximation. As expected,  $\avg{E_g}<\avg{E_m}$, so none of the
modes are energetically dominated by self-gravity.

%no GI for odd modes

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 2cm 0cm
    0cm]{figures/compare_growth3_tilted_even.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.52cm]{figures/compare_growth3_tilted_odd.ps} 
  \caption{Growth rates in massive isothermal disks with $Q=~0.2$ and
    $\beta=10$ for a range of azimuthal field strengths $B_y/B_z$. The
    dots are solutions computed from the full problem, with the
    colorbar measuring the gravitational potential perturbation, while
    the solid curves are computed from the Cowling approximation. The
    top panel correspond to even modes and the
    bottom panel correspond to odd modes, which have
    $W^\prime(0)=0$ and $W(0)=0$, respectively, for $\epsilon\to0$.
    \label{compare_growth3_tilted}}
\end{figure}

@@

Consider first the even modes in the top panel of 
Fig. \ref{compare_growth3_tilted}. %Self-gravity slightly stabilizes
%modes with small $k_x$ as $B_y$ is increased. 
Self-gravity destabilizes modes with  $k_xH\gtrsim O(1)$. Consequently, the
cut-off wavenumber is larger when SG is included. %get smaller scale
                                %MRI 
Destablization is most effective for purely vertical fields: with
$\epsilon=0,\, 
k_xH\simeq 1.4$, SG increases the growth rate by $\sim 30\%$. Although
these modes are fundamentally magnetic, this is consistent with
\cite{goldreich65a}, who find SG can only destabilize 
symmetric density perturbations, which is the case for $B_y=0$ here.  
With increasing $\epsilon$, we find the density perturbation $W$ deviates from an even 
function, $|W(0)|$ decreases and $\max{(|W|)}$ moves off of the mid-plane.  
%density pert at the midplane goes down
%We expect this to reduce the destabilization
%effect of SG,  
Together with the increased total magnetic pressure with 
$\epsilon$, destabilization by SG weakens. 
Thus, the Cowling approximation becomes increasingly good with stronger
$B_y$ for the even modes.

%This destabilization effect weakens with
%$B_y$ because the total magnetic pressure increases, which generally
%opposes GI, although the modes are fundamentally associated with MRI.

The modes in the bottom panel of Fig. \ref{compare_growth3_tilted} 
display opposite behavior. For $B_y=0$, $W$ is odd and 
self-gravity has no effect. When $B_y>0$, we find $W$ deviates from an odd
function and the mid-plane density perturbation $|W(0)|$ increases. 
SG is stabilizing for these modes at all wavelengths, and is most
effective at $k_xH =  O(1)$.  
%destabilizing effect is absent  
%for $\epsilon=0,\, k_xH\simeq 1.2$, SG decreases the growth rate by $\sim
%13\%$.

Fig. \ref{result_tilted} show eigenfunctions for $\epsilon=3$ and
$k_xH=1.2$ with and without the Cowling approximation. These differ
from the even modes considered above in that the horizontal magnetic
perturbations maximize at $z=0$ (cf. Fig. \ref{mri_massive_resis}).  
SG significantly enhances the mid-plane density perturbation, making the
gravitational potential energy comparable to the magnetic energy, 
which becomes more confined near the mid-plane.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
    0cm]{figures/result_tilted_cowling.ps}  
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
    0.cm]{figures/result_tilted_fullsg.ps} 
  \caption{Energy densities for a MRI mode in an isothermal ideal disk
    with an azimuthal field $B_y = 3B_z$, computed in the Cowling
    approximation (top) and with full self-gravity (bottom). These
    modes correspond to those in the bottom panel of
    Fig. \ref{compare_growth3_tilted}.    
    \label{result_tilted}}
\end{figure}

%strong field, compressibility 
To interpret the above result for odd modes, note 
that compressibility affects the MRI in the presence of an 
azimuthal field, even in a non-self-gravitating disk. If the perturbed 
disk remains in vertical hydrostatic equilibrium, then 
\begin{align} 
  |W|\sim \frac{B_y}{\mu_0\rho}|\dby|,
\end{align}
to order of magnitude in a non-SG disk, so a strong toroidal field can
cause a large density perturbation
\citep{pessah05}. Fig. \ref{result_tilted} shows that this is
amplified by self-gravity.

We checked that for the modes in Fig. \ref{result_tilted} the vertical
velocity is small, $|\dvz|/\left(|\dvx|^2+|\dvy|^2\right)^{1/2}
\lesssim 0.2$, so the above argument applies. Compressibilty is 
enhanced by both the azimuthal field and self-gravity, which tends
to stabilize the MRI \citep{kim00}. Its effect is important here  
because the azimuthal Alfven speed is sonic. The destabilization
effect of SG is absent (or unimportant) because the density
perturbation is not symmetric.

%The results for even and odd modes are do not contradict each
%other. Self-gravity tends to be destabilizing, but this requires the
%density perturbation to be symmetric, as is the case for the even
%modes. 
%For the MRI however, compressibility has a stabilizing effect

%scale separation for odd modes? because modes with kx=1 will be
%quenched

% for $\beta=10$ and
%$\epsilon=3$. 
%because magnetic energy is concentrated near midplane    
%magnetic pressure, need thermal pressure 
%This
%can subsequently be amplitfied by its self-gravity. 
%That is, compressibility is enhanced by both
%For $\epsilon=3$ and $\beta=10$ the azimuthal Alven speed is
%sonic. Compressibility becomes important for strong toroidal
%fields. If the system remains in approximately vertical hydrod

\subsection{Resistive disks with GI}
Here we examine an isothermal, resistive disk with $Q=0.18$ and
$\Lambda_0=0.1$. GI is permitted, and has a similar growth rate as the
MRI.

### 2014-03-31 paper/results_gi.tex
tex updates

\subsection{Co-existence of MRI and GI}
We first consider ideal MHD by setting $\Lambda_0=100$ and $A=1$. 
Fig. \ref{compare_growth3} show growth rates for 
modes with $k_xH=1$ as a function of $\beta$ in disks with $Q=0.18$,
$Q=0.14$ and $Q=0.12$. The colorbar quantifies the importance of
self-gravity with 
\begin{align}
 \tau \equiv \frac{\avg{E_g}}{\avg{E_g} + \avg{E_m}}. 
\end{align} 
Thus, modes with $\tau  = 1$ are
energetically dominated by self-gravity (GI) and $\tau\ll1$ represents a mode
dominated by magnetic perturbations (MRI).

@@

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity can enhance or reduce
MRI growth rates depending on $k_x$, but the effect is small.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth3_Q0d12_B20.ps}  
  \caption{Growth rates of unstable modes in the massive isothermal
    disk with $Q=0.12$ and $\beta=20$, as a function of the horizontal
    wavenumber $k_x$. The colorbar measures the importance of
    self-gravity. The solid line corresponds to MRI modes in the
    Cowling approximation. The dashed line corresponds to pure GI
    modes, obtained by including a high resisitivity in the full problem. 
    \label{compare_growth3_Q01d2}}
\end{figure}

>>>>

\subsection{Co-existence of MRI and GI}
We first consider ideal MHD by setting $\Lambda_0=100$ and $A=1$. 
Fig. \ref{compare_growth3} show growth rates for 
modes with $k_xH=1$ as a function of $\beta$ in disks with $Q=0.18$,
$Q=0.14$ and $Q=0.12$.

@@

Comparing pure MRI (solid line) and the magnetically-dominated
portions of $\gamma(k_x)$ show that self-gravity increases MRI growth
rates at large $k_x$, but the effect is small. Note that this
destabilization is through the linear response, rather than through
the background stratification.

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_growth3_Q0d12_B20.ps}  
  \caption{Growth rates of unstable modes in the massive isothermal
    disk with $Q=0.12$ and $\beta=20$, as a function of the horizontal
    wavenumber $k_x$. The colorbar measures the importance of
    self-gravity. The solid line corresponds to MRI modes in the
    Cowling approximation. The dashed line corresponds to pure GI
    modes, obtained by including a high resisitivity in the full
    problem. 
%    {\bf REDO FIGURE}
    \label{compare_growth3_Q01d2}}
\end{figure}

### 2014-03-31 paper/setup.tex
tex updates

\section{Local disk model}\label{setup}
We study the local stability of an inviscid, self-gravitating and
magnetized fluid disk orbiting a central mass with
potential $\Phi_*(r,z)$, where $(r,\varphi,z)$ are cylindrical
co-ordinates from the central mass. We use the shearing box approximation     
\citep{goldreich65b} and consider a small patch of the disk at
$r=r_0$. The local frame rotates at angular velocity 
$\Omega_0=\Omega(r_0,0)$ about the central mass, where $r\Omega^2 =
\p\Phi_*/\p r$. We also define $S=-r\p\Omega/\p r$ as the local shear
rate and $\Omega_z^2\equiv\p^2\Phi_*/\p z^2$ as the square of the
local vertical frequency.

@@

\subsection{Equilibrium disk} 
The unperturbed disk is steady and described by
$\rho=\rho(z)$, $\bm{B} = B_z\hat{\bm{z}} + B_y\hat{\bm{y}}$ where
$B_{y,z}$ are constants and the toroidal field strength is
$B_y=\epsilon B_z$. The equilibrium 
velocity field is $\bm{v} = -Sx\hat{\bm{y}}$. We consider Keplerian
disks so that $S = 3\Omega/2$ and the epicyclic frequency 
$\kappa\equiv\sqrt{2\Omega(2\Omega-S)}=\Omega=\Omega_z$. 
%$\Omega = \Omega_z.$ 
We assume a thin disk and neglect the radial component of the
self-gravitational force in 
the unperturbed disk.

The equilibrium density field is obtained by solving
\begin{align}
  &0=\frac{1}{\rho}\frac{d P}{dz} + \Omega_z^2z + \frac{d\Phi_d}{dz},\label{eqm_eqns1}\\
  &\frac{d^2\Phi_d}{dz^2} = 4\pi G \rho.\label{eqm_eqns2}
\end{align}
We consider \begin{inparaenum}[(i)]
\item isothermal disks with $P=c_{s0}^2\rho$\label{iso_eos}; 
\item polytropic disks with $P=K\rho^2$ with $K=c_{s0}^2/2\rho_0$;
\end{inparaenum}
where $\rho_0\equiv\rho(0)$ is the mid-plane density.
The sound speed $c_s\equiv\sqrt{dP/d\rho}$ 
so that $c_{s0}$ is the global sound speed in the isothermal disk, and
is the mid-plane sound speed in the polytropic disk. For the polytropic
disk the disk thickness $H$ is such that $\rho(H)=0$. Since the
isothermal disk has no surface, we define $H$ such that
$\rho(H)=\epsilon\rho(0)$ with $\epsilon=10^{-2}$. A non-dimensional
measure of the disk thickness is given by 
\begin{align}
  f^{-1} = \frac{H\Omega}{\csmid}, 
\end{align}
and $f$ will appear in subsequent discussions.

@@

\subsection{Resistivity profile}\label{resis_profile}
We consider constant resistivity or 
resistivity prescription such that $\eta(z)$ increases away from the
midplane. In the latter case, we follow \cite{fleming03} and adopt the
resistivity profile 
\begin{align}
  \eta(z) =
  \sqrt{2}\eta_0\left[\exp{\left(-g_+\right)}+\exp{\left(-g_-\right)}\right]^{-1/2},  
\end{align}
where
\begin{align}
  &g_\pm(z) =  \frac{\Sigma_\pm(z)-\Sigma_0}{\Sigma_*}, \\
  &\Sigma_\pm(z) = \int_{\pm z}^\infty\rho(z^\prime)dz^\prime, \label{sigma_pm}
\end{align}
and $\Sigma_0\equiv\Sigma_{\pm}(0)$, so that $g_\pm(0)=0$ and $\eta_0 
= \eta(0)$. The constant $\Sigma_*$ is chosen such that 
\begin{align}
  \cosh{\left(\frac{\Sigma_0}{\Sigma_*}\right)} =
  \left[\frac{\eta_0}{\eta(\infty)}\right]^2,
\end{align}
and we define $\eta_0/\eta(\infty)\equiv A$ as the conductivity
boost factor from the midplane to the disk surface. We remark that
once the equilibrium $\rho$ and $d\rho/dz$ are obtained from
Eq. \ref{eqm_eqns1} --- \ref{eqm_eqns2}, the integration for
Eq. \ref{sigma_pm} can be performed implicitly by using Poisson's 
equation.

We use the Elsasser number $\Lambda$ as a non-dimensional measure of
conductivity,
\begin{align} 
  \Lambda \equiv \frac{v_A^2}{\eta\Omega},
\end{align}
where $v_A \equiv B_z/\sqrt{\mu_0\rho}$ is the vertical Alfven speed. 
%Example
%profiles of $\Lambda$ are shown in Fig. \ref{eqm_resis} for the
%isothermal disk. 
The disk may be considered ideal where $\Lambda
\gtrsim 1$. 
%Since $\hat{\rho}(z/H)$ is weakly dependent on  self-gravity, the
%ideal region only changes slightly as $Q$ is decreased.

@@

\subsection{System parameters}
We summarize here the main parameters describing the resistive,
self-gravitating shearing box. The strength of self-gravity is
parametrized by 
\begin{align}
  Q \equiv \frac{\Omega^2}{4\pi G\rho_0}
\end{align}
\citep{mamat10}, which is used to set the mid-plane density $\rho_0$. 
The relation between $Q$ and the Toomre parameter for gravitational
instability of razor-thin disks is described in Appendix \ref{q3d2d}.

The plasma $\beta$ measures the inverse strength of the  
magnetic field, 
\begin{align}
  \beta \equiv \frac{\csmid^2}{v_{A0}^2} =
  \frac{\csmid^2\mu_0\rho_0}{B_z^2},  
\end{align}
where $v_{A0}$ is the midplane Alfven speed. The strength of
resistivity is measured by the midplane Elsasser number 
\begin{align}
  \Lambda_0 \equiv\Lambda(0) =  \frac{v_{A0}^2}{\eta_0\Omega}. %= \frac{f^2 R_m}{\beta}, 
\end{align}
%where the third equality defines the magnetic Reynolds number
%$R_m=H^2\Omega/\eta_0$ and
%where $f\equiv \csmid/H\Omega$ is a numerical
%factor of order unity. %measures thickness

>>>>

\section{Local disk model}\label{setup}
We study the local stability of an inviscid, self-gravitating and
magnetized fluid disk orbiting a central star with
potential $\Phi_*(r,z)$, where $(r,\varphi,z)$ are cylindrical
co-ordinates from the star. We use the shearing box approximation     
\citep{goldreich65b} to consider a small patch of the disk at
a fiducial radius $r=r_0$. The local frame rotates at angular velocity 
$\Omega_0=\Omega(r_0,0)$ about the star, where $r\Omega^2 =
\p\Phi_*/\p r$. We also define $S=-r\p\Omega/\p r$ as the local shear
rate and $\Omega_z^2\equiv\p^2\Phi_*/\p z^2$ as the square of the
local vertical frequency.

@@

\subsection{Equilibrium disk} 
The unperturbed disk is steady and described by
$\rho=~\rho(z)$, $\bm{B} = B_z\hat{\bm{z}} + B_y\hat{\bm{y}}$ where
$B_{y,z}$ are constants and the toroidal field strength is
$B_y=\epsilon B_z$. The equilibrium 
velocity field is $\bm{v} = -Sx\hat{\bm{y}}$. We consider Keplerian
disks so that $S = 3\Omega/2$ and the epicyclic frequency 
$\kappa\equiv\sqrt{2\Omega(2\Omega-S)}=\Omega=\Omega_z$. 
%$\Omega = \Omega_z.$ 
We assume a thin disk and neglect the radial component of the
self-gravitational force in 
the unperturbed disk.

The equilibrium density field is obtained by solving
\begin{align}
  &0=\frac{1}{\rho}\frac{d P}{dz} + \Omega_z^2z + \frac{d\Phi_d}{dz},\label{eqm_eqns1}\\
  &\frac{d^2\Phi_d}{dz^2} = 4\pi G \rho.\label{eqm_eqns2}
\end{align}
We consider \begin{inparaenum}[(i)]
\item isothermal disks with $P=c_{s0}^2\rho$\label{iso_eos}; 
\item polytropic disks with $P=K\rho^2$ with $K=c_{s0}^2/2\rho_0$;
\end{inparaenum}
where $\rho_0\equiv\rho(0)$ is the mid-plane density.
The sound speed $c_s\equiv\sqrt{dP/d\rho}$ 
so that $c_{s0}$ is the global sound speed in the isothermal disk, and
is the mid-plane sound speed in the polytropic disk. For the polytropic
disk the disk thickness $H$ is such that $\rho(H)=0$. Since the
isothermal disk has no surface, we define $H$ such that
$\rho(H)=10^{-2}\rho_0$. A non-dimensional
measure of the disk thickness is given by 
\begin{align}
  f^{-1} = \frac{H\Omega}{\csmid}, 
\end{align}
and $f$ will appear in subsequent discussions.

@@

\subsection{Resistivity profile}\label{resis_profile}
We consider constant resistivity or a 
resistivity prescription such that $\eta(z)$ increases away from the
mid-plane. In the latter case, we follow \cite{fleming03} and adopt the
resistivity profile 
\begin{align}
  \eta(z) =
  \sqrt{2}\eta_0\left[\exp{\left(-g_+\right)}+\exp{\left(-g_-\right)}\right]^{-1/2},  
\end{align}
where
\begin{align}
  &g_\pm(z) =  \frac{\Sigma_\pm(z)-\Sigma_0}{\Sigma_*}, \\
  &\Sigma_\pm(z) = \int_{\pm z}^\infty\rho(z^\prime)dz^\prime, \label{sigma_pm}
\end{align}
and $\Sigma_0\equiv\Sigma_{\pm}(0)$, so that $g_\pm(0)=0$ and $\eta_0 
= \eta(0)$. The constant $\Sigma_*$ is chosen such that 
\begin{align}
  \cosh{\left(\frac{\Sigma_0}{\Sigma_*}\right)} =
  \left[\frac{\eta_0}{\eta(\infty)}\right]^2,
\end{align}
and we define $\eta_0/\eta(\infty)\equiv A$ as the conductivity 
boost factor from the mid-plane to the disk surface. We remark that
once $\rho$ and $d\rho/dz$ are obtained from
Eq. \ref{eqm_eqns1} --- \ref{eqm_eqns2}, the integration for
Eq. \ref{sigma_pm} can be performed implicitly by using Poisson's 
equation.

We use the Elsasser number $\Lambda$ as a non-dimensional measure of
conductivity,
\begin{align} 
  \Lambda \equiv \frac{v_A^2}{\eta\Omega},
\end{align}
where $v_A \equiv B_z/\sqrt{\mu_0\rho}$ is the vertical Alfven speed. 
%Example
%profiles of $\Lambda$ are shown in Fig. \ref{eqm_resis} for the
%isothermal disk. 
Because of the density stratification, the Elsasser number 
increases with height even for constant resistivity. The disk may be
considered ideal where $\Lambda \gtrsim 1$. 
%Since $\hat{\rho}(z/H)$ is weakly dependent on  self-gravity, the
%ideal region only changes slightly as $Q$ is decreased.

@@

\subsection{System parameters}
The main parameters describing the resistive, self-gravitating
shearing box is as follows. The strength of self-gravity is
parametrized by 
\begin{align}
  Q \equiv \frac{\Omega^2}{4\pi G\rho_0}
\end{align}
\citep{mamat10}, which is used to set the mid-plane density $\rho_0$. 
The relation between $Q$ and the Toomre parameter for gravitational
instability of razor-thin disks, $Q_\mathrm{2D}$, is described in
Appendix \ref{q3d2d}.

The plasma $\beta$ measures the inverse strength of the  
magnetic field,  
\begin{align}
  \beta \equiv \frac{\csmid^2}{v_{A0}^2} =
  \frac{\csmid^2\mu_0\rho_0}{B_z^2},  
\end{align}
where $v_{A0}$ is the midplane Alfven speed. Note that we use 
the vertical field for this definition throughout the paper. The
strength of conductivity is measured by the midplane Elsasser number  
\begin{align}
  \Lambda_0 \equiv\Lambda(0) =  \frac{v_{A0}^2}{\eta_0\Omega}. %= \frac{f^2 R_m}{\beta}, 
\end{align}
%where the third equality defines the magnetic Reynolds number
%$R_m=H^2\Omega/\eta_0$ and
%where $f\equiv \csmid/H\Omega$ is a numerical
%factor of order unity. %measures thickness

### 2014-03-31 paper/summary.tex
tex updates

\section{Summary and discussion}\label{summary}
In this paper, we have examined the linear stability of radially
structured three-dimensional disks with non-uniform entropy
distribution. These calculations may be considered as an extension to 
the 2D Rossby wave instability \citep{li00} by adding the vertical
dimension, or to the barotropic RWI calculations of \citetalias{lin12} by
adding an energy equation with a simpler numerical method.

We adopted polytropic disk equilibria so that
the magnitude of entropy gradients can be conveniently parametrized
by $\Delta\gamma\equiv\gamma/\Gamma$, and we focused on the effect of
$\Delta\gamma\geq1$. When the background density and
velocity field is fixed through $\Gamma$, we found increasing $\Delta
\gamma$ has negligible effect on the instability growth rate. However,
{\bf the magnitude of} pressure and density perturbations increase with height, and the 
meridional flow associated with the vortex core is qualitatively
changed, with the introduction of {\bf meridional} vortical
motion.

{\bf Meridional vortical motion was found to correlate with a small 
  tilt of a fluid column with negative vertical vorticity
  perturbation.  In standard hydrodynamics, vorticity tilting
  can originate from a contribution of the
  $\bm{\omega}\cdot\nabla\bm{v}$ term in the evolution equation of the
  vorticity independently of the baroclinic source term $\nabla \rho
  \times\nabla p$. However, given the tilt is absent in our homentropic
  calculations, we associate the tilt with the baroclinic source term, which
  produces azimuthal vorticity. } 
We also found
that the vertical velocity at the vortex core is no longer linear in
$z$, as for homentropic flow.

In our second set of experiments, we fixed $\gamma$ and decreased
$\Gamma$. We found that by making the flow nonhomentropic, the
co-rotation region became \emph{more} three-dimensional, despite the
decrease in growth rate. This result
is opposite to \citetalias{lin12} where lowering $\Gamma$ made the
flow less three-dimensional. This implies that entropy gradients  
play an important role in the vertical structure of the perturbations. 
%This will likely have impact
%on the motion of dust particles within Rossby vortices, in particular 
%as the flow is dominated by bouyancy forces away from the midplane.

We also considered isothermal equilibria. A linear calculation with 
$\Gamma=1.1$ and one with a strictly isothermal setup ($\Gamma\equiv
1$) were consistent. Both produced prominent meridional 
vortical motion. In order to verify this feature, we ran a nonlinear
simulation of the RWI in an initially isothermal disk, but evolved
adiabatically. We indeed identified said  vortical motion.    
%although the agreement with linear calculations is
%qualitative. This might imply that nonlinear effects become important
%for the RWI vertical structure earlier than in the horizontal
%direction. 
Keeping in mind that the setup for linear and nonlinear simulations 
were not identical (e.g. numerical grid, boundary treatment), similarities
between them, such as mode frequency, growth rate and horizontal flow,
are satisfactory.

Vortical motion in the meridional plane thus appears characteristic of 
the \emph{linear} RWI in nonhomentropic disks. Whether or not this is 
significant for the vortex evolution can only be answered by detailed 
long term nonlinear simulations. If this vortical motion is present 
in the nonlinear regime then it may prevent dust particles from reaching the disk surface, 
which occurs for homentropic flow \citep{meheut12c}.

However, given this meridional vortical motion is absent in the homentropic 
linear solution, it may eventually vanish because of entropy
mixing, if no mechanism is present to maintain entropy gradients. 
For example, the background entropy increases with height but 
the linear entropy perturbation becomes more negative with height, {\bf 
and its magnitude grows exponentially in time}. 
{\bf Indeed,  recent 3D fully compressible simulations in
  nonhomentropic disks shows that well into the nonlinear regime,
  Rossby vortices are columnar
  \citep{richard13}}. On the other hand, \cite{meheut12} observed
strong meridional vortical motion in their homentropic hydrodynamic
simulations; we conclude they are of nonlinear origin.

In the linear solutions, we often observe perturbation magnitudes
increase away from the midplane in nonhomentropic
disks\footnote{This reminds us of the off-midplane vortices
  discovered by \cite{barranco05} in nonlinear local simulations, 
%  which suggest the non-zero vertical bouyancy frequency to be
%  stabilizing   against secondary instabilities \citep{lesur09}. 
  but the  
  setup considered in that study is very different from the
  present work. Nevertheless, in both cases the vertical entropy
  gradient is stabilizing away from the midplane. 
  %%  Nevertheless, a positive vertical bouyancy frequency
  %% may stabilize 3D vortices \citep{lesur09}.
} 
(e.g. Fig. \ref{lin_nonlin_rz}). Then the RWI may not be as robust
against vertical boundary conditions as it is to radial boundary
conditions. This could pose difficulty for the RWI to develop in dead
zones of real protoplanetary disks, which are expected to be confined
from above and below by magnetically turbulent layers
\citep{oishi09}. The vertical boundary condition set by these layers
may or may not be compatible with the linear RWI solution.

\subsection{Caveats and outlooks}
%need a good guess
One trade-off for the simplicity of our numerical method for 
linear simulations is that a trial eigenfrequency must be
guessed. This is not a significant obstacle for the problem at
hand, because previous RWI studies provide an important guide
\citep[][]{li00}. Otherwise, zeros of the complex function
$\mathcal{D}(\sigma)=\mathrm{det}\,\bm{U}$ need to be located with more 
rigorous methods \citep[e.g.][]{kojima86,valborro07}. We have also exploited
previous findings that the PPI and RWI are predominantly two-dimensional 
\citep{papaloizou85, goldreich86, kojima89, umurhan10, meheut12, lin12, lin12c}, which
enabled the use of a small number of basis functions. 
However, there could exist parameter regimes where the RWI
has significant vertical structure, rendering our solution method
inefficient.

%extension to baroclinic equilibria 
%simple polytrope background
Our conclusions are limited to polytropic backgrounds. While this was 
convenient for numerical experiments, it is an over-simplification
of protoplanetary disks, which are 
expected to have complicated vertical structure \citep{terquem08}. In
particular, we found that entropy gradients plays a role in the
vertical structure of the {\bf linear} RWI, and even a small entropy gradient can noticeably
modify the vertical flow (\S\ref{nz_nonzero}).
Thus, a realistic model for entropy evolution is needed. 
%This is important because even a small entropy gradient can noticeably 
%modify the vertical flow (\S\ref{nz_nonzero}).

It would also be of interest to generalize the calculations
to \emph{baroclinic} equilibria\footnote{In fact, baroclinic tori were
  briefly considered by \cite{frank88}.}, 
for which $\p_z\Omega\neq0$. This may well be the 
case when the equilibrium pressure depends on both the density and temperature. 
%This requires 
%derivation of the linearized equations including terms involving $\p_z\Omega$. 
Complications from baroclinic instabilities may arise, however \citep{knobloch86, umurhan12, nelson12}.
%% This may involve adding heating or cooling
%% sources to the energy equation. 
%source terms - friction, cooling
%new physics - self-gravity

We have neglected gas self-gravity in this study. Our models therefore
assume that the Toomre parameter is much larger than unity in both the 
unperturbed \emph{and} perturbed states. {\bf However, self-gravity
  may affect the RWI even when the Toomre parameter is not small
  \citep{lovelace12}}. 
Previous studies
have found  higher $m$ RWI modes are favored when disk self-gravity is
included \citep{lyra08,lin11a}. Recent 3D simulations of the RWI in a
locally isothermal disk show that vertical self-gravity can noticeably
enhance the density perturbation near the midplane,
even though the initial disk was considered low mass \citep{lin12b}. %%  The effects
%%  of self-gravity %% should be considered in conjunction with disk
%%  thermodynamics. %In a nonlinear simulation, the latter may not hold even if the former
%does.
%Previous studies of the RWI also find higher $m$ modes modes are
%favored when disk self-gravity is included
%\citep{lyra08,lin11a}.

In principle, one can express the Poisson integral as a matrix
operator and  incorporate it into our formalism. {\bf The linear} problem is
further complicated by the need of a numerical solution to the
equilibrium equations describing a radially structured,
self-gravitating 3D disk \citep{muto11}.  Such a linear calculation is
beyond the scope of this paper, but will be inevitable for
understanding  the RWI in 3D self-gravitating disks. Perhaps a simpler
starting point, to gain first insight, is  direct hydrodynamic 
simulations {\bf including disk gravity}. This is indeed the approach taken in our follow-up paper.    
%\clearpage
%since the ZEUS-MP code already includes a Poisson solver (as used in Lin 2012b).

%have shown that
%self-gravity reduces the RWI growth rate through the potential perturbation, 
%but may be destabilizing through modification of the background disk \citep{lin11a}.

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we have performed explicit stability calculations of
magnetized, self-gravitating disks in the local approximation.

%layered accretion incompatable in massive disks

%why did previous MRI simulations find no difference between SG and
%nonSG? prob because the mri mode has too small vertical wavelength

\subsection{Caveats and outlooks}
We discuss below two major extensions to our linear model that
should be undertaken, before performing non-linear simulations of
magnetized, self-gravitating disks, which is our eventual goal.

\emph{Beyond the shearing box.} The shearing box ignores the curvature
of toroidal field lines present in the global disk
geometry. \cite{pessah05} have shown that new effects on the MRI arise
when the curvature of a super-thermal toroidal field is accounted for;
although \citeauthor{pessah05} focused on modes with large (small)
vertical (radial) wavenumbers, for which we do not expect self-gravity
to be important. Since compressibilty becomes important for strong
toroidal fields, the effect of self-gravity may become even more
significant when super-thermal toroidal fields are
considered. However, global disk models will be neccessary
to self-consistently probe this regime.

%
%This
%prevents consideration of super-thermal toroidal fields
%\citep{pessah05}.

%tension stabilizes GI  
%This prevents us from considering super-thermal
%toroidal fields.  
%go to cylindrical disk geometry (fu & lai) 
%tension stabilizes GI

%spiral arms, GI angular mometum distribution 
\emph{Beyond axisymmetry.} It is well known that self-gravitating
disks can be unstable to non-axisymmetric perturbations but stable
against axisymmetric perturbations, although the threshold Toomre
parameter for both is still of order unity. %papaloizou/lin,
                                %sajvoini??
